Generating Random Logic Programs Using Constraint Programming

### 2020-05-16 text/paper2/paper.tex
Edit the section on custom constraints

\documentclass[runningheads]{llncs}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}
\usepackage{blkarray}
\usepackage[binary-units]{siunitx}
\usepackage[inline]{enumitem}
\usepackage{capt-of}
\usepackage[caption=false]{subfig}

@@

@@

% TODO: explain about entailment and propagation

A ProbLog program has to be stratified in order to ensure a unique
solution to every query
\cite{DBLP:journals/jlp/BalbinPRM91,DBLP:conf/padl/MantadelisR17}. Independence
is a fundamental concept in probability theory that imposes structure on the
probability distribution and makes probabilistic inference easier. In the
context of probabilistic logic programming languages, these two seemingly
disparate concepts can be defined using the same building block, i.e., a
predicate dependency graph. In this section, we define both notions and describe
custom constraints meant to ensure that our generated programs are valid and
have the desired independence structure. A new constraint is usually presented
with two algorithms: one for \emph{entailment} checking, i.e., checking if the
(partially solved) constraint model satisfies the constraint or not, and one for
\emph{propagation}, i.e., removing values from domains that are incompatible
with the current state of the model.

Let $\mathscr{P}$ be a probabilistic logic program. Its \emph{predicate
  dependency graph} is a directed graph $G_{\mathscr{P}} = (V, E)$ with the set
of nodes $V$ consisting of all predicates in $\mathscr{P}$. For any two
different predicates $\mathsf{P}$ and $\mathsf{Q}$, we add an edge from
$\mathsf{P}$ to $\mathsf{Q}$ if there is a clause in $\mathscr{P}$ with
$\mathsf{Q}$ as the head and $\mathsf{P}$ mentioned in the body. We say that the
edge is \emph{negative} if there exists a clause with $\mathsf{Q}$ as the head
and at least one instance of $\mathsf{P}$ at the body such that the path from
the root to the $\mathsf{P}$ node in the tree representation of the clause
passes through at least one negation node. Otherwise, it is \emph{positive}. We
say that $\mathscr{P}$ (or $G_{\mathscr{P}}$) has a \emph{negative cycle} if
$G_{\mathscr{P}}$ has a cycle with at least one negative edge. A program
$\mathscr{P}$ is \emph{stratified} if $G_{\mathscr{P}}$ has no negative cycles.
Thus a simple entailment algorithm for stratification can be constructed by
selecting all clauses, all predicates of which are fully determined, and looking
for negative cycles in the dependency graph constructed based on those clauses
using an algorithm such as Bellman-Ford.

Let $\mathsf{P}$ be a predicate in a program $\mathscr{P}$. The set of
\emph{dependencies} of $\mathsf{P}$ is the smallest set $D_{\mathsf{P}}$ such
that $\mathsf{P} \in D_{\mathsf{P}}$, and, for every $\mathsf{Q} \in
D_{\mathsf{P}}$, all direct predecessors of $\mathsf{Q}$ in $G_{\mathscr{P}}$
are in $D_{\mathsf{P}}$. Two predicates $\mathsf{P}$ and $\mathsf{Q}$ are
\emph{independent} if $D_{\mathsf{P}} \cap D_{\mathsf{Q}} = \emptyset$.

\begin{example} \label{ex:program}
  Consider the following (fragment of a) program:
  \begin{align}
    \mathsf{sibling}(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z), \nonumber \\
    \mathsf{father}(X, Y) &\gets \mathsf{parent}(X, Y) \land \neg\mathsf{mother}(X, Y). \label[clause]{eq:example_clause}
  \end{align}
  Its predicate dependency graph is in \cref{fig:predicate_dependencies}.
  Because of the negation in \eqref{eq:example_clause} (as seen in
  \cref{fig:example_tree2}), the edge from $\mathsf{mother}$ to
  $\mathsf{father}$ is negative, while the other two edges are positive.
  The dependencies of each predicate are:
  \begin{alignat*}{3}
    D_{\mathsf{parent}} &= \{ \mathsf{parent} \}, \quad && D_{\mathsf{sibling}}
    &&= \{\mathsf{sibling}, \mathsf{parent} \},\\
    D_{\mathsf{mother}} &= \{ \mathsf{mother} \}, \quad && D_{\mathsf{father}}
    &&= \{ \mathsf{father}, \mathsf{mother}, \mathsf{parent} \}.
  \end{alignat*}
  Hence, we have two pairs of independent predicates, i.e., $\mathsf{mother}$ is
  independent of $\mathsf{parent}$ and $\mathsf{sibling}$.
\end{example}
\begin{figure}[t]
  \centering
  \begin{minipage}{.49\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw,circle,gray,text=black] (and) at (0, 0) {$\land$};
      \node[draw,circle,gray,text=black] (not) at (1.5, -1) {$\neg$};
      \node[draw,gray,text=black] (parent) at (-1.5, -1) {$\mathsf{parent}(X, Y)$};
      \node[draw,gray,text=black] (mother) at (1.5, -2) {$\mathsf{mother}(X, Y)$};
      \draw[gray] (and) -- (parent);
      \draw[gray] (and) -- (not);
      \draw[gray] (not) -- (mother);
    \end{tikzpicture}%
    \captionof{figure}{A tree representation of the body of \eqref{eq:example_clause}}
    \label{fig:example_tree2}
  \end{minipage}
  \begin{minipage}{.49\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
      \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
      \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
      \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (sibling);
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (father);
      \draw[-{Stealth[scale=1.5]}] (mother) edge node[below] {$-$} (father);
    \end{tikzpicture}%
    \captionof{figure}{The predicate dependency graph of the program in
      \cref{ex:program}. Positive edges are labelled with `$+$', and negative
      edges with `$-$'.}
    \label{fig:predicate_dependencies}
  \end{minipage}
\end{figure}

Next, we can define a $|\predicates{}| \times |\predicates{}|$ adjacency matrix
$\mathbf{A}$ with a binary domain (i.e., without positivity/negativity). It can
be defined element-wise by stating that $\mathbf{A}[i][j] = 0$ if and only if,
for all $k \in \{ 0, \dots, \maxNumClauses{} - 1 \}$, either
$\variable{heads}[k].\variable{predicate} \ne j$ or $ i \not\in
\{a.\variable{name} \mid a \in \variable{bodies}[k].\variable{values} \}$.

Given an undetermined model, we can classify all dependencies of a predicate
$\mathsf{P}$ into three categories based on how many of the edges on the path
from the dependency to $\mathsf{P}$ are undetermined. In the case of zero, we
call the dependency \emph{determined}. In the case of one, we call it
\emph{almost determined}. Otherwise, it is \emph{undetermined}. In the context
of propagation and entailment algorithms, we define a \emph{dependency} as the
sum type: \syntax{<dependency> ::= $\Determined(p)$ | $\Undetermined(p)$ |
  $\AlmostDetermined(p, s, t)$}. Each alternative represents a determined,
undetermined, and almost determined dependency, respectively. Here, $p \in
\predicates{}$ is the name of the predicate which is the dependency of
$\mathsf{P}$, and---in the case of $\AlmostDetermined$---$(s, t) \in
\predicates{}^2$ is the one undetermined edge in $\mathbf{A}$ that prevents the
dependency from being determined. For a dependency $d$---regardless of its exact
type---we will refer to its predicate $p$ as $d.\mathsf{predicate}$. In
describing the algorithms, we will use an underscore to replace any of $p$, $s$,
$t$ in situations where the name is unimportant.

@@

\begin{algorithm}[t]
  \LinesNumbered
  \SetKwFunction{getDependencies}{deps}
  \SetKwFunction{fail}{fail}
  \SetKwFunction{removeValue}{removeValue}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \KwData{predicates $p_1$, $p_2$; adjacency matrix $\mathbf{A}$}
  \For{$(d_1, d_2) \in \getDependencies{$p_1$, 0} \times \getDependencies{$p_2$,
      0}$ s.t. $d_1.\predicate = d_2.\predicate$}{
    \lIf{$d_1$ {\bf is} $\Determined(\_)$ {\bf and} $d_2$ {\bf is}
      $\Determined(\_)$}{\fail{}} \label{line:fail}
    \If{$d_1$ {\bf is} $\Determined(\_)$ {\bf and} $d_2$ {\bf is}
      $\AlmostDetermined(\_, s, t)$ {\bf or} $d_2$ {\bf is} $\Determined(\_)$
      {\bf and} $d_1$ {\bf is} $\AlmostDetermined(\_, s,
      t)$ \label{line:propagation_condition}}{
      $\mathbf{A}[s][t]$.\removeValue{$1$}\; \label{line:propagation_result}
    }
  }
  \caption{Propagation for independence}
  \label{alg:independence_propagation}
\end{algorithm}

Propagation algorithms have two goals: causing a contradiction (failing) in
situations where the corresponding entailment algorithm would return
\textsc{false}, and eliminating values from domains of variables that are
guaranteed to cause a contradiction. \cref{alg:independence_propagation} does
the former on \cref{line:fail}. Furthermore, for any dependency shared between
predicates $p_1$ and $p_2$, if it is determined ($\Determined$) for one
predicate and almost determined ($\AlmostDetermined$) for another, then the edge
that prevents the $\AlmostDetermined$ from becoming a $\Determined$ cannot
exist--\cref{line:propagation_condition,line:propagation_result} handle this
possibility.

@@

>>>>

\documentclass[runningheads]{llncs}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{tikz}
\usepackage{mathrsfs}
\usepackage[nounderscore]{syntax}
\usepackage{blkarray}
\usepackage[binary-units]{siunitx}
\usepackage[inline]{enumitem}
\usepackage{capt-of}
\usepackage[caption=false]{subfig}
\usepackage{booktabs}

@@

TODO: Review \cite{DBLP:journals/tplp/WangWM15,DBLP:conf/ijcai/AmendolaRT17,DBLP:journals/ai/AmendolaRT20}

@@

In this section, we present custom constraints for stratification and
independence. Stratification is a condition that both logic programs and
probabilistic logic programs need to satisfy to ensure a unique answer
to every query
\cite{DBLP:journals/jlp/BalbinPRM91,DBLP:conf/padl/MantadelisR17}, while
independence is a fundamental concept in probability theory that imposes
structure on the probability distribution and simplifies inference. In the
context of probabilistic logic programming languages, these two seemingly
disparate concepts can be defined using the same building block, i.e., a
predicate dependency graph. Constraints are usually presented with two
algorithms: one for \emph{entailment} checking, i.e., checking if the (partially
solved) constraint model satisfies the constraint or not, and one for
\emph{propagation}, i.e., removing values from domains that are incompatible
with the current state of the model. We present both algorithms for independence
while settling for a short description of entailment for stratification.

Let $\mathscr{P}$ be a probabilistic logic program with predicates
$\predicates{}$. Its \emph{(predicate) dependency graph} is a directed graph
$G_{\mathscr{P}}$ with $\predicates{}$ as nodes and an edge between $\mathsf{P},
\mathsf{Q} \in \predicates{}$ if there is a clause in $\mathscr{P}$ with
$\mathsf{Q}$ as the head and $\mathsf{P}$ mentioned in the body. We say that the
edge is \emph{negative} if there exists a clause with $\mathsf{Q}$ as the head
and at least one instance of $\mathsf{P}$ at the body such that the path from
the root to the $\mathsf{P}$ node in the tree representation of the clause
passes through at least one negation node; otherwise, it is \emph{positive}. We
say that $\mathscr{P}$ (or $G_{\mathscr{P}}$) has a \emph{negative cycle} if
$G_{\mathscr{P}}$ has a cycle with at least one negative edge. A program
$\mathscr{P}$ is \emph{stratified} if $G_{\mathscr{P}}$ has no negative cycles.
Thus a simple entailment algorithm for stratification can be constructed by
selecting all clauses, all predicates of which are fully determined, and looking
for negative cycles in the dependency graph constructed based on those clauses
using an algorithm such as Bellman-Ford.

For any predicate $\mathsf{P} \in \predicates{}$, the set of \emph{dependencies}
of $\mathsf{P}$ is the smallest set $D_{\mathsf{P}}$ such that $\mathsf{P} \in
D_{\mathsf{P}}$, and, for every $\mathsf{Q} \in D_{\mathsf{P}}$, all direct
predecessors of $\mathsf{Q}$ in $G_{\mathscr{P}}$ are in $D_{\mathsf{P}}$. Two
predicates $\mathsf{P}$ and $\mathsf{Q}$ are \emph{independent} if
$D_{\mathsf{P}} \cap D_{\mathsf{Q}} = \emptyset$.

\begin{example} \label{ex:program}
  Consider the following (fragment of a) program:
  \begin{align}
    \mathsf{sibling}(X, Y) &\gets \mathsf{parent}(X, Z) \land \mathsf{parent}(Y, Z), \nonumber \\
    \mathsf{father}(X, Y) &\gets \mathsf{parent}(X, Y) \land \neg\mathsf{mother}(X, Y). \label[clause]{eq:example_clause}
  \end{align}
  Its predicate dependency graph is in \cref{fig:predicate_dependencies}.
  Because of the negation in \eqref{eq:example_clause}, the edge from
  $\mathsf{mother}$ to $\mathsf{father}$ is negative, while the other two edges
  are positive. The dependencies of each predicate are:
  \begin{alignat*}{3}
    D_{\mathsf{parent}} &= \{ \mathsf{parent} \}, \quad && D_{\mathsf{sibling}}
    &&= \{\mathsf{sibling}, \mathsf{parent} \},\\
    D_{\mathsf{mother}} &= \{ \mathsf{mother} \}, \quad && D_{\mathsf{father}}
    &&= \{ \mathsf{father}, \mathsf{mother}, \mathsf{parent} \}.
  \end{alignat*}
  Hence, we have two pairs of independent predicates, i.e., $\mathsf{mother}$ is
  independent of $\mathsf{parent}$ and $\mathsf{sibling}$.
\end{example}

\begin{figure}[t]
  \centering
  \begin{minipage}{.49\textwidth}
    \centering
    \begin{tikzpicture}
      \node[draw] (parent) at (0, 0.5) {$\mathsf{parent}$};
      \node[draw] (mother) at (0, -0.5) {$\mathsf{mother}$};
      \node[draw] (sibling) at (2, 0.5) {$\mathsf{sibling}$};
      \node[draw] (father) at (2, -0.5) {$\mathsf{father}$};
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (sibling);
      \draw[-{Stealth[scale=1.5]}] (parent) edge node[above] {$+$} (father);
      \draw[-{Stealth[scale=1.5]}] (mother) edge node[below] {$-$} (father);
    \end{tikzpicture}%
    \captionof{figure}{The predicate dependency graph of the program in
      \cref{ex:program}. Positive edges are labelled with `$+$', and negative
      edges with `$-$'.}
    \label{fig:predicate_dependencies}
  \end{minipage}
  \begin{minipage}{.49\textwidth}
    \centering
    \captionof{table}{Types of (potential) dependencies of a predicate
      $\mathsf{P}$ based on the number of undetermined edges on the path from
      the dependency to $\mathsf{P}$}
    \label{tbl:dependencies}
    \begin{tabular}{clc}
      \toprule
      Edges & Name & Notation \\
      \midrule
      $0$ & determined & $\Determined(p)$ \\
      $1$ & almost determined & $\AlmostDetermined(p, s, t)$ \\
      $>1$ & undetermined & $\Undetermined(p)$ \\
      \bottomrule
    \end{tabular}
  \end{minipage}
\end{figure}

Since the definition of independence relies on the dependency graph, we can
represent this graph as an adjacency matrix constructed as part of the model.
Let $\mathbf{A}$ be a $|\predicates{}| \times |\predicates{}|$ binary matrix
defined element-wise by stating that $\mathbf{A}[i][j] = 0$ if and only if, for
all $k = 0, \dots, \maxNumClauses{} - 1$, either
$\variable{heads}[k].\variable{predicate} \ne j$ or $ i \not\in
\{a.\variable{name} \mid a \in \variable{bodies}[k].\variable{values} \}$.

Given a partially-solved model with its predicate dependency graph, let us pick
an arbitrary path from $\mathsf{Q}$ to $\mathsf{P}$ (for some $\mathsf{P},
\mathsf{Q} \in \predicates{}$) that consists of determined edges that are
denoted by $1$ in $\mathbf{A}$ and potential/undetermined edges that are
denoted by $\{ 0, 1 \}$. Each such path characterises a \emph{(potential)
  dependency} $\mathsf{Q}$ for $\mathsf{P}$. We can put each such dependency
into one out of three classes depending on the number of undetermined edges on
the path. These classes are outlined in \cref{tbl:dependencies}, where $p$
represents the dependency predicate $\mathsf{Q}$, and, in the case of
$\AlmostDetermined{}$, $(s, t) \in \predicates{}^2$ is the one undetermined edge
on the path. For a dependency $d$---regardless of its exact type---we will refer
to its predicate $p$ as $d.\mathsf{predicate}$. In describing the algorithms, we
will use an underscore to replace any of $p$, $s$, $t$ in situations where the
name is unimportant.

@@

\begin{algorithm}[t]
  \LinesNumbered
  \SetKwFunction{getDependencies}{deps}
  \SetKwFunction{fail}{fail}
  \SetKwFunction{removeValue}{removeValue}
  \SetKwData{predicate}{predicate}
  \SetKwData{source}{source}
  \SetKwData{target}{target}
  \KwData{predicates $p_1$, $p_2$; adjacency matrix $\mathbf{A}$}
  \For{$(d_1, d_2) \in \getDependencies{$p_1$, 0} \times \getDependencies{$p_2$,
      0}$ s.t. $d_1.\predicate = d_2.\predicate$}{
    \lIf{$d_1$ {\bf is} $\Determined(\_)$ {\bf and} $d_2$ {\bf is}
      $\Determined(\_)$}{\fail{}} \label{line:fail}
    \lIf{$\{ d_1, d_2 \} = \{ \Determined(\_), \AlmostDetermined(\_, s, t)
      \}$ \label{line:propagation_condition}}{$\mathbf{A}[s][t]$.\removeValue{$1$}}
  }
  \caption{Propagation for independence}
  \label{alg:independence_propagation}
\end{algorithm}

Propagation algorithms have two goals: causing a contradiction (failing) in
situations where the corresponding entailment algorithm would return
\textsc{false}, and eliminating values from domains of variables that are
guaranteed to cause a contradiction. \cref{alg:independence_propagation} does
the former on \cref{line:fail}. Furthermore, for any dependency shared between
predicates $p_1$ and $p_2$, if it is determined ($\Determined$) for one
predicate and almost determined ($\AlmostDetermined$) for another, then the edge
that prevents the $\AlmostDetermined$ from becoming a $\Determined$ cannot
exist---\cref{line:propagation_condition} handles this possibility.

@@

%    \begin{tikzpicture}
%      \node[draw,circle,gray,text=black] (and) at (0, 0) {$\land$};
%      \node[draw,circle,gray,text=black] (not) at (1.5, -1) {$\neg$};
%      \node[draw,gray,text=black] (parent) at (-1.5, -1) {$\mathsf{parent}(X, Y)$};
%      \node[draw,gray,text=black] (mother) at (1.5, -2) {$\mathsf{mother}(X, Y)$};
%      \draw[gray] (and) -- (parent);
%      \draw[gray] (and) -- (not);
%      \draw[gray] (not) -- (mother);
%    \end{tikzpicture}%
%    \captionof{figure}{A tree representation of the body of \eqref{eq:example_clause}}
%    \label{fig:example_tree2}
